Descriptive, not injunctive, social norms caused increases in mask wearing throughout the COVID-19 pandemic

### 2022-10-13 manuscript.Rmd
Update manuscript.Rmd
first draft manuscript revision

abstract: |
 Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people _ought_ to do, whereas descriptive social norms inform what people _actually_ do. While previous experimental work has revealed people’s sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging a year and a half of longitudinal data from a representative sample of adults in the United States (15 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior are tightly coupled and both change quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.

note              : "This working paper has not yet been peer-reviewed. This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation."
keywords          : "descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "4826 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : apa.csl
urlcolor          : blue
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
tar_load_everything()
# to ensure that the Rmd runs last, load appendix files again
tar_load(c(d, plot1, plot2, plot4, plot5, plot6, plot7,
           m1.1, m1.2, m3.1, itemTable, lavaanTable,
           changePointsTable, constructValDes, constructValInj))
```

Social norms are a key aspect of human sociality. Broadly defined as behavioral guidelines enforced by groups of people without the law [@VanKleef2019], social norms have attracted the attention of psychologists for decades. Since Asch's [-@Asch1956] early studies of normative conformity, evidence has accrued that humans are particularly attuned to social norms [@Tomasello2014]. From a young age, children begin to adhere to and enforce group-wide social norms [@Jensen2014; @Rakoczy2013; @Schmidt2012] and both children and adults rely on normative emotions, such as shame and guilt, to determine when they or others have violated social norms [@Vaish2011; @Schaumberg2022]. This uniquely human sensitivity to social norms allows groups of unrelated people to cooperate and coordinate in the face of existential threats, such as resource scarcity, natural disasters, and infectious diseases [@Gelfand2011; @Henrich2015].

Previous research has distinguished between two primary kinds of social norms: injunctive norms and descriptive norms [@Cialdini1991; @Deutsch1955]. Injunctive social norms are societal, indicating what others in the group tend to approve or disapprove of, and often involve social sanctions if violated. By contrast, descriptive norms are situational, simply describing what most people are doing in a given situation. Though these two kinds of social norms tend to align, they can also be in conflict with one another. For example, there may be an injunctive norm that cleaning up litter at a picnic site is the right thing to do: one _ought_ to behave this way. However, if an individual observes that most people are leaving their litter behind at the site, the descriptive norm is to not clean up. It is thus possible for injunctive and descriptive norms to have independent effects on behavior [@Schultz2007].

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], social norms have remained an elusive concept in the behavioral sciences and several open questions remain [@Fehr2018a; @VanKleef2019]. In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms influence behavior?

@@

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the behavioral impact of social norms. For example, field experiments have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on injunctive norms [@Schultz2007]. However, previous studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers, and thus do not allow cooperative social norms to emerge and affect behavior naturally over time within a population. An alternative way to identify causality, whilst retaining ecological validity, is to follow perceptions of social norms and cooperative behavior over time amidst a real, unfolding social dilemma. This is particularly important because social norms are not static: they change dynamically over time through processes of deliberation and social interaction [@Smith2015].

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, we studied mask wearing behavior during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, two months into the pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a measure that people should take to minimize the spread of COVID-19. But mask wearing posed a social dilemma to individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the evolution of mask wearing in the United States throughout the COVID-19 pandemic allows us to study, on a short timescale within a single population, the emergence of novel injunctive and descriptive social norms and the causal effects of these norms on cooperative behavior.

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between between-person and within-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic.

Here, we use a year and a half of longitudinal data from a representative sample of adults in the United States (15 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. We aimed to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms affect mask wearing behavior over time?

@@

Using the platform Prolific (https://www.prolific.co/), we distributed surveys to a representative sample of adults from the United States (_n_ = `r nrow(d)`, _M_~age~ = `r round(mean(d$Age.1, na.rm = TRUE))` years, `r round(mean(d$Ethnicity.1 == "White or Caucasian", na.rm = TRUE)*100)`% White, `r round(mean(d$Sex.1 == 2, na.rm = TRUE)*100)`% Women; see Supplementary Figure \@ref(fig:plotUSMap) for geographic distribution). From September 2020 to February 2022, we asked participants to complete regular surveys of COVID-19 related attitudes and behaviors. This resulted in 15 unique time points of data collection throughout the pandemic. The first 12 time points were distributed monthly, while the last three time points were distributed every two months. `r sum(!is.na(d$StartDate.2))` of the initial `r nrow(d)` participants returned to complete the survey at Time 2, while `r sum(!is.na(d$StartDate.15))` participants continued through to Time 15 (see Supplementary Figure \@ref(fig:plotAttrition) for attrition rates across all time points).

@@

In 8 of the 15 time points (Time 2, 3, 5, 9, 11, 13, 14, and 15), we asked questions about perceived descriptive and injunctive mask wearing norms.

@@

To account for potential confounding in our longitudinal analysis, we constructed a directed acyclic causal graph outlining the expected causal relationships between our variables (see Supplementary Figure \@ref(fig:plotDAG)). In this causal model, we included two additional variables that could plausibly act as common causes of perceived social norms and mask wearing behavior: county-level COVID-19 prevalence and county-level political leaning. If COVID-19 is prevalent in a participant’s county, then there might be more mask wearing overall, from people in participants’ local area, respected individuals, and from participants themselves, causing increases in both self-reported mask wearing behavior and perceived social norms. Similarly, left-leaning counties might have increased levels of self-reported mask wearing behavior and perceived social norms, as the political context affects the proportion of individuals following COVID-19 protective measures [@BaxterKing2022]. It is therefore important to statistically control for these confounding variables.

We retrieved data on county-level COVID-19 prevalence from a CDC public health surveillance dataset [@CDC2022a]. In the CDC dataset, COVID-19 prevalence was operationalized as the total new COVID-19 cases per 100,000 population in counties over the past 7 days. Since this dataset was updated regularly starting from February 2022, and so each county had multiple entries, we averaged and logged this variable within each county to get the log average weekly new COVID-19 cases per 100,000 population. We retrieved data on county-level political leaning from the MIT Election Lab [@MITElectionLab]. For each county, the dataset stated how many votes the Republican, Democratic, Green Party, and other candidates received in the 2020 Presidential election. We used this information to determine which candidate won the vote in that county.

>>>>

abstract: |
 Social norms allow humans to coordinate and cooperate in the face of existential threats. In particular, injunctive social norms prescribe what people _ought_ to do, whereas descriptive social norms inform what people _actually_ do. While previous experimental work has revealed people’s sensitivity to normative influence, several open questions remain about the natural emergence of injunctive and descriptive social norms within populations and their influences on cooperative behavior over time. To understand how social norms emerge and shape behavior in a non-experimental setting, we studied mask wearing during the COVID-19 pandemic. Leveraging two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = `r nrow(tar_read(d))`), we tracked people’s reported mask wearing behavior and their perceived injunctive and descriptive mask wearing norms as the pandemic unfolded. Longitudinal trends of norm perceptions and self-reported mask wearing suggested that norms and behavior are tightly coupled and both change quickly in response to recommendations from public health authorities. In addition, a random-intercept cross-lagged panel model revealed that descriptive norms, but not injunctive norms, caused future increases in mask wearing behavior. These findings underline the relative importance of descriptive social norms in shaping cooperative behavior during uncertain times.

note              : "This working paper has not yet been peer-reviewed. This study was funded by the Interdisciplinary Cooperation Initiative, ASU President's Office, the Cooperation Science Network, the Institute for Mental Health Research, the University of New Mexico, the Indiana University College of Arts & Sciences, the Rutgers University Center for Human Evolutionary Studies, the Charles Koch Foundation, and the John Templeton Foundation."
keywords          : "descriptive norms; injunctive norms; longitudinal; COVID-19; mask wearing; cooperation"
wordcount         : "xxxx words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
tar_load_everything()
# to ensure that the Rmd runs last, load appendix files again
tar_load(c(d, plot1, plot2, plot4, plot5, plot6,
           m1.1, m1.2, itemTable, lavaanTable,
           changePointsTable, constructValDes, constructValInj))
```

Social norms are a key aspect of human sociality. Broadly defined as commonly known behavioral guidelines enforced by groups of people, social norms have attracted the attention of behavioral scientists for decades [@Bicchieri2009; @Cialdini1991; @Fehr2018a; @Legros2020; @VanKleef2019]. Since Asch's early studies of normative conformity [@Asch1956], evidence has accrued that humans are particularly attuned to social norms [@Tomasello2014]. From a young age, children begin to adhere to and enforce group-wide social norms [@Jensen2014; @Rakoczy2013; @Schmidt2012] and both children and adults rely on normative emotions, such as shame and guilt, to determine when they or others have violated social norms [@Vaish2011; @Schaumberg2022]. This uniquely human sensitivity to social norms allows groups of unrelated people to cooperate and coordinate in the face of existential threats, such as resource scarcity, natural disasters, and infectious diseases [@Gelfand2011; @Henrich2015].

Previous research has distinguished between two primary kinds of social norms: injunctive norms and descriptive norms [@Bicchieri2009; @Cialdini1991; @Deutsch1955]. Injunctive social norms indicate what others in the group tend to approve or disapprove of and often involve social sanctions if violated. By contrast, descriptive norms are situational, simply describing what most people are doing in a given situation. Though these two kinds of social norms tend to align, they can also be in conflict with one another. For example, there may be an injunctive norm that cleaning up litter at a picnic site is the right thing to do: one _ought_ to behave this way. However, if an individual observes that most people are leaving their litter behind at the site, the descriptive norm is to not clean up. It is thus possible for injunctive and descriptive norms to have independent effects on behavior [@Schultz2007].

Despite decades of research on the causes and consequences of injunctive and descriptive norms [@Bicchieri2009; @Cialdini1991; @Cialdini2007; @Cialdini2021; @Schultz2007], social norms have remained an elusive concept in the behavioral sciences and several open questions remain [@Fehr2018a; @VanKleef2019]. In the current work, we focus on two such questions. First, how do injunctive and descriptive norms emerge naturally over time within a population? Second, how do evolving injunctive and descriptive norms influence behavior?

@@

With regards to normative influences on behavior, there is a wealth of cross-sectional evidence demonstrating the behavioral impact of social norms. For example, field experiments have demonstrated the positive effects of descriptive norms on a variety of cooperative behaviors, including recycling [@Nigbur2010], paying taxes [@Larkin2018], and sustainably reusing towels in hotels [@Goldstein2008]. Evidence also suggests that any potentially deleterious effects of descriptive social norms (e.g., choosing to litter at a picnic site that already contains visible signs of littering) can be counteracted by instead focusing individuals’ attention on injunctive norms [@Schultz2007]. However, these cross-sectional studies have two main limitations. First, studies have not adequately controlled for other potential non-social influences on behavior, such as factual beliefs and personal normative beliefs [@Bicchieri2014]. For example, descriptive social norms might only affect cooperative behavior indirectly by influencing perceptions of the effectiveness of the behavior (i.e., factual beliefs) or individuals' moral preferences (i.e., personal normative beliefs). Second, cross-sectional studies have tended to follow experimental designs in which perceptions of social norms are manipulated by the researchers at a single time point, and thus do not allow cooperative social norms to emerge and affect behavior naturally over time within a population. An alternative way to identify causality, whilst retaining ecological validity, is to follow perceptions of social norms and cooperative behavior over time amidst a real, unfolding social dilemma. This is particularly important because social norms are not static: they change dynamically over time through processes of deliberation and social interaction [@Smith2015].

To understand how novel injunctive and descriptive social norms emerge over time and shape cooperative behavior in a non-experimental setting, we studied mask wearing behavior during the COVID-19 pandemic. Before the pandemic, mask wearing was not a common behavior in the United States. In April 2020, two months into the pandemic, mask wearing was officially recommended by the Centers for Disease Control and Prevention (CDC) as a cooperative behavior that people should adopt to minimize the spread of COVID-19. But mask wearing posed a social dilemma to individuals, in that it imposed personal costs (e.g., difficulty breathing, disrupted social interaction) for the benefit of the wider community (e.g., "flattening the curve" to protect at-risk individuals). Thus, the evolution of mask wearing in the United States throughout the COVID-19 pandemic allows us to study, on a short timescale within a single population, the emergence of novel injunctive and descriptive social norms and the causal effects of these norms on cooperative behavior.

Recent research has found positive relationships between perceptions of social norms and protective COVID-19 behaviors. In the United States, one cross-sectional study found that perceptions of injunctive norms positively predicted intentions to stay at home to minimize exposure [@Macy2021], and another vignette study found that experimentally-manipulated descriptive norms increased personal mask wearing intentions [@Bokemper2021]. In Germany, a two-wave study found that perceptions of descriptive norms positively predicted future protective behaviors, such as physical distancing [@Rudert2021]. These studies are telling, but since they are experimental, cross-sectional, or only minimally longitudinal, they are unable to distinguish between between-person and within-person change over time [@Hamaker2015], nor do they have the temporal granularity to capture fluctuating changes in norm strength and norm adherence across the entire pandemic. These studies also do not control for the influences of non-social beliefs, such as factual beliefs and personal normative beliefs [@Bicchieri2014].

Here, we use two years of longitudinal data from a representative sample of adults in the United States (18 time points; _n_ = 916) to track the development of descriptive and injunctive mask wearing norms and mask wearing behavior over the course of the COVID-19 pandemic. We aimed to answer two main research questions. First, how do descriptive and injunctive mask wearing norms emerge and evolve over time in the United States population? Second, how do descriptive and injunctive mask wearing norms affect mask wearing behavior over time?

(ref:plotTimelineCaption) _Timeline of self-reported mask wearing behavior and perceived social norms in the United States throughout the COVID-19 pandemic._ (a) Points and line ranges indicate mean averages and standard errors for the self-reported mask wearing item. This item was measured across all eighteen time points on a 5-point Likert scale, with higher values indicating increased frequency of personal mask wearing during in-person interactions. (b) Points and line ranges indicate mean averages and standard errors for perceived descriptive mask wearing norms (blue) and perceived injunctive mask wearing norms (green). These items were measured across eleven time points on a 7-point Likert scale, with higher values indicating stronger perceived social norms. (c) Smoothed data for new COVID-19 cases in the United States, displayed on the log scale (data retrieved from Our World in Data; https://ourworldindata.org/). Across all panels, gray dashed lines represent significant pandemic-related events in the United States, such as vaccine approval from the Food and Drug Administration (FDA) and public health recommendations from the Centers for Disease Control and Prevention (CDC).

```{r plotTimeline, echo=F, fig.width=6, fig.height=6, fig.cap="(ref:plotTimelineCaption)"}
plot3
```

Two main observations can be made about the emergence and stability of social norms from these visualizations. First, social norms and behavior were tightly coupled over time. Although social norms are measured on fewer occasions than mask wearing behavior, we can see that as mask wearing behavior decreased in the summer of 2021, so too did perceived descriptive and injunctive mask wearing norms. Subsequently, the steep rise in COVID-19 case numbers in the fall of 2021 saw concomitant increases in both mask wearing behavior and perceived social norms, though to lower levels than before. In line with these patterns, multilevel regression models revealed positive correlations between mask wearing behavior and perceived descriptive mask wearing norms (_b_ = `r printnum(fixef(m1.1)["DesNorms"])`, 95% confidence interval [`r printnum(conf1.1["DesNorms","2.5 %"])` `r printnum(conf1.1["DesNorms","97.5 %"])`]) and between mask wearing behavior and perceived injunctive mask wearing norms (_b_ = `r printnum(fixef(m1.2)["InjNorms"])`, 95% CI [`r printnum(conf1.2["InjNorms","2.5 %"])` `r printnum(conf1.2["InjNorms","97.5 %"])`]) across individuals and time points (Supplementary Figure \@ref(fig:plotCorBehNorm); Supplementary Table \@ref(tab:modelSummaryTable1)).

Second, fluctuations in mask wearing behavior and perceived social norms are in line with recommendations broadcasted by the CDC, the main national public health agency of the United States. We do not have data for the very start of the pandemic in early 2020, but the high levels of mask wearing and strong perceived social norms at the start of our observation window likely emerged after the initial mask wearing recommendation from the CDC in April 2020. Perceived social norms and mask wearing behavior subsequently declined after the CDC rescinded their mask wearing recommendation in March 2021, and then increased again after the CDC updated their guidelines for indoor mask use in high risk areas in August 2021. Finally, perceived social norms and mask wearing declined again after the CDC eased restrictions in March 2022. Impressively, shifts in mask wearing behavior and norms happened on the timescale of weeks. These trends were confirmed by a series of multilevel regression models with change points aligning with changes in CDC mask wearing recommendations (Supplementary Figure \@ref(fig:plotCDCSens); Supplementary Table \@ref(tab:changePointsTable)).

Sample averages can provide informative trends, but they do not allow us to estimate within-person changes in mask wearing behavior and perceived social norms over time. To determine whether within-person changes in social norms temporally preceded within-person changes in mask wearing behavior, we fitted a ten-wave unconstrained random-intercept cross-lagged panel model to the longitudinal data. This structural equation model separately estimated stable trait-like between-person individual differences and within-person fluctuations from trait levels for our main variables and controls: perceived descriptive mask wearing norms, perceived injunctive mask wearing norms, factual beliefs, personal normative beliefs, and self-reported mask wearing behavior (see Supplementary Figure \@ref(fig:plotDAG) for justification of control variables). According to established fit statistics, this model fitted the data well (root mean square error of approximation = `r printnum(fitMeasures['rmsea'], digits = 3)`, 95% CI [`r printnum(fitMeasures['rmsea.ci.lower'], digits = 3)` `r printnum(fitMeasures['rmsea.ci.upper'], digits = 3)`]; standardized root mean squared residual = `r printnum(fitMeasures['srmr'], digits = 3)`; comparative fit index = `r printnum(fitMeasures['cfi'], digits = 3)`). Since we are primarily interested in the causal effects of social norms on behavior, we focus on these variables in what follows (but see Supplementary Table \@ref(tab:lavaanTable) for full list of estimated autoregressive and cross-lagged effects).

```{r, echo = FALSE}
sc <- standardizedSolution(riclpm)
pe <- parameterEstimates(riclpm)
```

Regarding between-person individual differences, the covariances between the random intercepts in the model revealed positive correlations between stable trait levels of mask wearing behavior and perceived social norms. On average across the whole study, participants who more frequently wore masks during in-person interactions also perceived stronger descriptive mask wearing norms (_r_ = `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri3","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri3","ci.lower"])` `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri3","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "ri1" & pe$op == "~~" & pe$rhs == "ri3","pvalue"])`) and stronger injunctive mask wearing norms (_r_ = `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri2","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri2","ci.lower"])` `r printnum(sc[sc$lhs == "ri1" & sc$op == "~~" & sc$rhs == "ri2","ci.upper"])`], _p_ `r printp(pe[pe$lhs == "ri1" & pe$op == "~~" & pe$rhs == "ri2","pvalue"])`). Stable trait perceptions of descriptive and injunctive mask wearing norms were also highly positively correlated (_r_ = `r printnum(sc[sc$lhs == "ri2" & sc$op == "~~" & sc$rhs == "ri3","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "ri2" & sc$op == "~~" & sc$rhs == "ri3","ci.lower"])` `r printnum(sc[sc$lhs == "ri2" & sc$op == "~~" & sc$rhs == "ri3","ci.upper"])`], _p_ `r printp(pe[pe$lhs == "ri2" & pe$op == "~~" & pe$rhs == "ri3","pvalue"])`).

Regarding within-person dynamics over time, Figure \@ref(fig:plotRICLPM) displays autoregressive and cross-lagged effects for perceived descriptive norms, perceived injunctive norms, and mask wearing behavior across the study duration, controlling for non-social beliefs and political orientation. In random intercept cross-lagged panel models, autoregressive effects represent "persistence" or "inertia" in within-person fluctuations from stable trait levels. In other words, a positive autoregressive effect indicates that being higher than average on one measure predicts being higher than average on that same measure in the following time point (this is not to be confused with the "stability" of the measure over time, which is captured by the random intercepts in our model). By contrast, and most relevant for the current study, cross-lagged effects represent the effect of a within-person fluctuation in one measure on future within-person fluctuations in other measures. In other words, a positive cross-lagged effect indicates that being higher than average on one measure predicts being higher than average on _another_ measure in the following time point. Cross-lagged effects are thus used to infer within-person causal influences over time.

(ref:plotRICLPMCaption) _Results of ten-wave unconstrained random-intercept cross-lagged panel model._ Arrows represent within-person autoregressive and cross-lagged effects from the model, partitioning out stable between-person individual differences and controlling for factual beliefs, personal normative beliefs, and political orientation. Arrow thickness is scaled according to standardized effect size. Bolded arrows indicate significantly positive parameters, _p_ < 0.05. Gray arrows indicate non-significant parameters.

```{r plotRICLPM, echo=F, fig.width=7.5, fig.height=3.5, fig.cap="(ref:plotRICLPMCaption)"}
plot7
```

In late 2020 and throughout 2021, we see several cross-lagged effects from perceived descriptive norms to future mask wearing behavior. On four occasions, within-person increases in perceived descriptive norms predicted future within-person increases in mask wearing behavior. According to recent effect size guidelines for cross-lagged panel models [@Orth2022], the standardized beta coefficients for these cross-lagged effects were large (Time 2, $\beta$ = `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.lower"])` `r printnum(sc[sc$lhs == "w1_05" & sc$op == "~" & sc$rhs == "w3_02","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_05" & pe$op == "~" & pe$rhs == "w3_02","pvalue"])`; Time 5, $\beta$ = `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.lower"])` `r printnum(sc[sc$lhs == "w1_09" & sc$op == "~" & sc$rhs == "w3_05","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_09" & pe$op == "~" & pe$rhs == "w3_05","pvalue"])`; Time 11, $\beta$ = `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.lower"])` `r printnum(sc[sc$lhs == "w1_13" & sc$op == "~" & sc$rhs == "w3_11","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_13" & pe$op == "~" & pe$rhs == "w3_11","pvalue"])`; Time 13, $\beta$ = `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","est.std"])`, 95% CI [`r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.lower"])` `r printnum(sc[sc$lhs == "w1_14" & sc$op == "~" & sc$rhs == "w3_13","ci.upper"])`], _p_ = `r printp(pe[pe$lhs == "w1_14" & pe$op == "~" & pe$rhs == "w3_13","pvalue"])`). These cross-lagged effects from descriptive norms to mask wearing diminish in 2022. We also find some evidence for a reciprocal effect, whereby within-person increases in mask wearing behavior predicted future within-person increases in perceived descriptive norms. Moreover, several cross-lagged effects emerged between perceived descriptive and injunctive norms, demonstrating reciprocal within-person causal effects between these variables.

However, the model reveals that, after controlling for perceived descriptive norms, non-social beliefs, and political orientation, within-person changes in perceived injunctive norms did not predict future within-person changes in mask wearing behavior across the entire pandemic. All cross-lagged effects from perceived injunctive norms to mask wearing behavior are non-significant. Any causal effect that perceived injunctive norms might have had on future mask wearing behavior appears to be fully mediated by perceived descriptive norms. For example, between August 2021 and December 2021, perceived injunctive norms predicted future perceived descriptive norms, which themselves predicted future mask wearing behavior. But aside from these indirect effects, perceived injunctive norms did not have a direct causal effect on mask wearing behavior over time within individuals.

Using longitudinal data from the United States across the entire COVID-19 pandemic, we aimed to understand how descriptive and injunctive mask wearing norms emerge and influence behavior in response to a naturally unfolding social dilemma. The trends of norm perceptions and self-reported mask wearing over time suggest that norms and behavior are tightly coupled, and both change quickly in response to recommendations from public health authorities. The results of our structural equation model also indicate that descriptive norms caused future increases in mask wearing behavior at the beginning of the pandemic. By contrast, injunctive norms were not directly causally related to future mask wearing behavior over the course of the pandemic.

In our longitudinal analysis, we found that _descriptive_ norms, not injunctive norms, independently predicted future increases in mask wearing. In line with this finding, descriptive norms have also been shown to predict future increases in physical distancing and prosocial behaviors (e.g., neighborhood help, charitable donations) throughout the COVID-19 pandemic, though this previous work did not adequately disentangle between-person and within-person effects [@Rudert2021]. Similarly, experimental work has revealed that people in the United States are more likely to report intentions to wear a mask if they are told that others are wearing masks [@Bokemper2021]. Why this specific effect of descriptive norms on mask wearing? Descriptive norms are particularly useful for fast changing, threatening situations with a high degree of uncertainty, such as the COVID-19 pandemic [@Gelfand2015]. During times of uncertainty, people look to others to quickly coordinate their behavior, and attempt to alleviate uncertainty-related stress by identifying with their group and its social norms [@Hogg2008; @Wellen1998]. Supporting this situational-uncertainty explanation, our model revealed that descriptive norms predicted future mask wearing at the beginning of the pandemic, when uncertainty was likely at its height. Unsure of how to respond to a novel threatening situation, people began to look to their neighbors and adapted their mask wearing behavior accordingly. However, as the uncertainty surrounding the pandemic lessened in 2022, reliance on descriptive norms appeared to decrease.

Our finding that injunctive norms do not predict future mask wearing behavior is at odds with cross-sectional evidence showing that perceived injunctive norms are positively correlated with intentions to stay indoors during the pandemic among older adults [@Macy2021]. One possible explanation for these conflicting findings is that, unlike previous work, our model systematically controlled for non-social beliefs, such as factual beliefs and personal normative beliefs, which could potentially have driven previous correlations. Another explanation is that, due to the increased opportunities to observe mask wearing in public, descriptive norms of mask wearing behavior were made more salient than injunctive norms throughout the pandemic, and therefore had a greater influence on behavior [@Cialdini1991]. By contrast, for more private behaviors like remaining indoors, it would have been less possible to observe other people’s behaviors, increasing the relative salience of injunctive norms. To test this idea, future research should expand our longitudinal cross-lagged approach to protective behaviors beyond mask wearing, including both public behaviors (e.g., physical distancing) and private behaviors (e.g., hand washing and home isolation).

@@

Using the platform Prolific (https://www.prolific.co/), we distributed surveys to a representative sample of adults from the United States (_n_ = `r nrow(d)`, _M_~age~ = `r round(mean(d$Age.1, na.rm = TRUE))` years, `r round(mean(d$Ethnicity.1 == "White or Caucasian", na.rm = TRUE)*100)`% White, `r round(mean(d$Sex.1 == 2, na.rm = TRUE)*100)`% Women; see Supplementary Figure \@ref(fig:plotUSMap) for geographic distribution). From September 2020 to October 2022, we asked participants to complete regular surveys of COVID-19 related attitudes and behaviors. This resulted in 18 unique time points of data collection throughout the pandemic. The first 12 time points were distributed monthly, while the remaining six time points were distributed every two months. `r sum(!is.na(d$StartDate.2))` of the initial `r nrow(d)` participants returned to complete the survey at Time 2, while `r sum(!is.na(d$StartDate.18))` participants continued through to Time 18 (see Supplementary Figure \@ref(fig:plotAttrition) for attrition rates across all time points).

@@

In 11 of the 18 time points (Time 2, 3, 5, 9, 11, 13, 14, 15, 16, 17, and 18), we asked questions about perceived descriptive and injunctive mask wearing norms.

@@

To identify direct causal effects in our longitudinal analysis, we constructed a directed acyclic causal graph outlining the expected causal relationships between our variables (see Supplementary Figure \@ref(fig:plotDAG)). In this causal model, we included two kinds of non-social beliefs highlighted by previous research [@Bicchieri2014]: factual beliefs (i.e., beliefs about the effectiveness or consequences of mask wearing) and personal normative beliefs (i.e., personal beliefs about whether mask wearing is the morally right thing to do). These variables were included as potential mediators of the effects of descriptive and injunctive social norms on mask wearing behavior. In addition, we also included political orientation as a common cause of all other variables. This is justified by evidence showing that mask wearing was heavily politicized in the United States throughout the pandemic [@BaxterKing2022]. Given this causal graph, it is necessary to control for factual beliefs, personal normative beliefs, and political orientation in order to estimate the direct causal effects of descriptive and injunctive norms on mask wearing behavior over time.
